```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
out.width = "75%"
```

Loading the dataset containing wheat seed measurements.

```{r seeds}
seeds<-read.csv("seeds.csv")
```

## Data Validation

```{r clean, echo=FALSE}
# Verify that the dataset does not contain missing values.
print(colSums(is.na(seeds)))
print(any(is.null(seeds)))

# Check whether the dataset contains negative or invalid numeric values.
print(sum(rowMeans(seeds < 0), na.rm = TRUE))
# Check for duplicated observations in the dataset.
print(dim(unique(seeds))[1])

```
The dataset contains no missing values in any of the variables and includes 210 complete cases.
In addition, no invalid or spurious values were identified in the dataset.
No negative values were detected across the variables.
The dataset also contains no duplicated records.

The following section provides an overview of the dataset structure and feature distribution.
```{r print, echo=FALSE}
str(seeds)
summary(seeds)

```
The dataset consists of 199 wheat grain samples belonging to three different varieties.
Seven quantitative variables related to the internal kernel structure, extracted using a soft X-ray imaging technique, are included in the dataset.
The target variable represents the wheat variety, while the remaining features describe area, perimeter, compactness, kernel length, kernel width, asymmetry coefficient, and kernel groove length.
```{r visualiz_data,echo=FALSE}
library(corrplot)
## corrplot 0.92 loaded
M = cor(seeds)
corrplot(M, method = 'number') # colorful number
# Perform PCA on the dataset.
prcomp_seeds <- prcomp(seeds[, 1:7])
pca_seeds <- data.frame(
  PC1 = prcomp_seeds$x[, 1],
  PC2 = prcomp_seeds$x[, 2],
  classification = seeds$Type
)
# Load visualization library for PCA projection.
library(ggplot2)
ggplot(pca_seeds, aes(x = PC1, y = PC2, col = classification)) +
  geom_point()
```
As shown in the correlation matrix, there is a very strong correlation of 0.99 between area and perimeter.
A moderate correlation of 0.61 can be observed between area and compactness, while the lowest correlation appears between perimeter and compactness with a value of 0.53.
Additional noticeable relationships include the correlations between area and kernel length, as well as between perimeter and kernel length.
These observations suggest that grain size related attributes have a meaningful influence on the dataset structure.

In this section, Min-Max normalization is applied to scale the feature values into a range between 0 and 1.

```{r normilize,echo=FALSE}
normalize <- function(x) {
return ((x - min(x)) / (max(x) - min(x)))
}

n_data <- as.data.frame(lapply(seeds[1:8], normalize))


summary(n_data)
```
he normalization process transforms all feature values into a consistent scale ranging from 0 to 1.

The resulting train_sample object contains 160 randomly selected observations.
```{r train_test_sets,echo=FALSE}
set.seed(1234)
shuffled_data= seeds[sample(1:nrow(seeds)), ]


train_sample <- sample(199, 160)

# Split the dataset into training and testing subsets.
data_train <- shuffled_data[train_sample, ]
data_test <- shuffled_data[-train_sample, ]

# Extract labels for each subset.
data_train_labals <- shuffled_data[train_sample, 8]
data_test_labals <- shuffled_data[-train_sample, 8]

# Verify class distribution across the subsets.
prop.table(table(data_train$Type))
prop.table(table(data_test$Type))
```
The dataset split preserves approximately 30% defaulted loans representation across both subsets.

The following section presents the implementation and evaluation of the classification models used in this project.

## K-Nearest Neighbors (KNN)
The KNN algorithm was evaluated using three different values of `k`: 3, 5, and 7.
```{r knn,echo=FALSE}
library(class) 
library(gmodels)
# Apply the KNN algorithm using different k values.
seeds_test_pred_k5 <- knn(train = data_train, test = data_test, cl = data_train_labals, k=5)
seeds_test_pred_k3 <- knn(train = data_train, test = data_test, cl = data_train_labals, k=3)
seeds_test_pred_k7 <- knn(train = data_train, test = data_test, cl = data_train_labals, k=7)

ACC.5 <- 100 * sum(data_test_labals == seeds_test_pred_k5)/NROW(data_test_labals) # For knn = 5
ACC.3 <- 100 * sum(data_test_labals == seeds_test_pred_k3)/NROW(data_test_labals)  # For knn = 3
ACC.7 <- 100 * sum(data_test_labals == seeds_test_pred_k7)/NROW(data_test_labals)  # For knn = 7
paste("Accuracy for k = 3",ACC.3,sep=" : ")
paste("Accuracy for k = 5",ACC.5,sep=" : ")
paste("Accuracy for k = 7",ACC.7,sep=" : ")
```

Among the evaluated configurations, the KNN model with k = 3 achieved the highest accuracy.

```{r knn_visoualize,echo=FALSE}
# Display the prediction confusion matrix
CrossTable(x = data_test_labals, y = seeds_test_pred_k3, prop.chisq=FALSE)
# Visualize KNN predictions using PCA projection.
prcomp_seeds <- prcomp(data_test)
pca_seeds <- data.frame(
  PC1 = prcomp_seeds$x[, 1],
  PC2 = prcomp_seeds$x[, 2],
  classification = seeds_test_pred_k3
)
# Load visualization library for PCA plotting.
ggplot(pca_seeds, aes(x = PC1, y = PC2, col = classification)) +
  geom_point() + # Show dots
  geom_text(
    label= (data_test_labals == seeds_test_pred_k3), 
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T
  )
```

## Decision Tree Model
The following section presents the implementation of a decision tree classifier using the C5.0 algorithm.

```{r dection_tree,echo=FALSE}
library(C50)

vars <- c("Area", "Perimeter","Compactness","Kernel.Length","Kernel.Width","Asymmetry.Coeff","Kernel.Groove")
seeds_tree_model <- C5.0(x=data_train[vars],y=as.factor(data_train$Type),trials = 3)
summary(seeds_tree_model)
plot(seeds_tree_model)
```
The trained decision tree model is evaluated on the test dataset.

```{r test_tree,echo=FALSE}
# Apply the trained model on the test dataset.
seeds_pred <- predict(seeds_tree_model, data_test)
CrossTable(data_test_labals, seeds_pred, prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE, dnn = c('actual Type', 'predicted Type'))

plot(seeds_pred)


ACC.withoutCost <- 100 * sum(data_test_labals == seeds_pred)/NROW(data_test_labals)

paste("Accuracy for DT without cost metrix",ACC.withoutCost,sep=" : ")

```
## Decision Tree with Cost Matrix
A cost matrix was introduced in order to penalize specific classification errors differently.
```{r cost,echo=FALSE}
matrix_dimensions <- list(c("1", "2", "3"), c("1", "2", "3"))
names(matrix_dimensions) <- c('actual Type', 'predicted Type')

error_cost <- matrix(c(0, 8, 10, 8, 0, 1, 1, 10, 0), nrow = 3, dimnames = matrix_dimensions)
error_cost
# Train the model using the cost matrix.
seeds_cost <- C5.0(data_train[vars], as.factor(data_train$Type), costs = error_cost)
plot(seeds_cost)
# Generate predictions on the test dataset.
seeds_cost_pred <- predict(seeds_cost, data_test)

CrossTable(as.factor(data_test$Type), seeds_cost_pred, prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE, dnn = c('actual Type', 'predicted Type'))

ACC.withoutCost <- 100 * sum(data_test_labals == seeds_pred)/NROW(data_test_labals)
ACC.cost <- 100 * sum(data_test_labals == seeds_cost_pred)/NROW(data_test_labals)

paste("Accuracy for DT without cost metrix",ACC.withoutCost,sep=" : ")
paste("Accuracy for DT with cost metrix",ACC.cost,sep=" : ")

```

The cost matrix configuration did not improve the model performance, as the overall accuracy remained unchanged.

The experimental results show that the KNN classifier achieved the strongest performance, reaching 100% prediction accuracy on the test dataset.
In comparison, the decision tree model achieved an accuracy of approximately 87%.

One possible explanation is the relatively small dataset size, where a limited number of misclassified samples has a substantial impact on the overall accuracy.

